---
title: "State Drug Utilization, 2000 - 2020"
author: 
  - Michael Quinn Maguire, MS ^[University of Florida, College of Pharmacy, Pharmaceutical Outcomes and Policy, michaelqmaguire2@cop.ufl.edu] 
  - Amie Goodin, PhD, MPP ^[University of Florida, College of Pharmacy, Pharmaceutical Outcomes and Policy, amie.goodin@cop.ufl.edu] 
  - Juan Hincapie-Castillo, PharmD, PhD, MS ^[University of Florida, College of Pharmacy, Pharmaceutical Outcomes and Policy, j.hincapie@cop.ufl.edu]
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

The following document provides a glimpse into the compiled drug utilization datasets from [Centers for Medicare & Medicaid Services](https://www.medicaid.gov/medicaid/prescription-drugs/state-drug-utilization-data/index.html).

New data extracts are pulled whenever the script 02_create-examine-date-metadata.R detects that there are differences between the previously extracted metadata and the current metadata on the website.

Data were compiled from the "full" datasets available in the dropdown menu. Overall national totals denoted with a state field of "XX" were removed after compiling the datasets into a single object.

Data were then matched to [IBM's Micromedex Red Book](https://www.ibm.com/products/micromedex-red-book) by the NDC code.

These plots show that there are some inconsistencies and data reporting issues for certain years. Please be cognizant of these issues when conducting your analyses or submitting your requests.

I produce more specialized plots when a request is made. If you think there is another type of plot that would be useful for this report, please contact me via email.

Please refer to the code in other areas of my [GitHub Repository](https://github.com/michaelqmaguire/medicaid-drug-utilization) for more specific information on how cleaning was performed. If you have any questions, please feel free to contact me at <michaelqmaguire2@cop.ufl.edu>.

Datasets are not publicly available due to size considerations. They can be made available upon request and discussion with my supervisors.

Datasets were imported and cleaned using R software. This report was written and compiled using RMarkdown. More information about RMarkdown can be located [here](https://rmarkdown.rstudio.com/).

### Setup

#### Loading libraries used to plot information

```{r libload, warning = FALSE, message = FALSE}
library(data.table)
library(hrbrthemes)
library(tidyverse)
library(tidylog)
```

#### Reading in the final dataset

```{r sdudload, include = TRUE}
sdudFinal <-
  fread(
    "./data/clean/03_sdud-redbook-final.csv",
    colClasses = c("proper_ndc" = "character")
  )

sdudFinal
```

### Creating aggregate files

#### Examine total row counts and summed number of prescriptions for entire dataset.

```{r sdudFinalAgg}
sdudFinal[
  i = ,
  j = .(
    RowCount = .N,
    TotalRx = sum(numberrx)
  )
]
```

#### Examine total row counts and summed number of prescriptions by year.

```{r sdudAggYear}
sdudAggYear <- 
  sdudFinal[
    i  = ,
    j  = .(
      RowCount = .N,
      TotalRx  = sum(numberrx, na.rm = TRUE),
      FakeDate = lubridate::ymd(year, truncated = 2L)
    ),
    by = year
  ][order(year)]

sdudAggYear
```

#### Examine total row counts and summed number of prescriptions by year and quarter.

```{r sdudAggQtrYear}
sdudAggQtrYear <- 
  sdudFinal[
    i  = ,
    j  = .(
      RowCount = .N,
      TotalRx  = sum(numberrx, na.rm = TRUE),
      FakeDate = lubridate::ymd(year, truncated = 2L)
    ),
    by = c("year", "quarter")
  ][order(year, quarter)]

sdudAggQtrYear
```

#### Examine total row counts and summed number of prescriptions by year and state.

```{r sdudAggStateYear}
sdudAggStateYear <-
  sdudFinal[
    i  = ,
    j  = .(
      RowCount = .N,
      TotalRx  = sum(numberrx, na.rm = TRUE),
      FakeDate = lubridate::ymd(year, truncated = 2L)
    ),
    by = c("year", "state")
  ]

sdudAggStateYear
```

#### Examine total row counts and summed number of prescriptions by year, state, and quarter.

```{r sdudAggQtrStateYear}
sdudAggQtrStateYear <-
  sdudFinal[
    i  = ,
    j  = .(
      RowCount = .N,
      TotalRx  = sum(numberrx, na.rm = TRUE),
      FakeDate = lubridate::ymd(year, truncated = 2L)
    ),
    by = c("quarter", "year", "state")
  ]

sdudAggQtrStateYear
```

### Create plots

#### Plot 1: Record count by year.

```{r plotYearRecordCount, fig.width = 20, fig.height = 12}
sdudAggYear %>%
  ggplot() +
    geom_col(
      aes(
        x    = FakeDate,
        y    = RowCount,
        fill = FakeDate
      ),
      color = "black"
    ) +
    scale_x_date(
      date_breaks = "1 year",
      date_labels = "%Y",
      expand      = c(.01, .01)
    ) +
    scale_y_continuous(labels = scales::comma) +
    scale_fill_viridis_c() + 
    theme_ipsum_rc(grid = "XY") +
    xlab("Year") +
    ylab("Number of Records") +
    labs(caption = "Source: https://www.medicaid.gov/medicaid/prescription-drugs/state-drug-utilization-data/index.html") +
    ggtitle("Number of Records by Year - 2000 to 2020") +
    theme(
      axis.title.x    = element_text(color = "black"),
      axis.text.x     = element_text(color = "black", angle = 90, vjust = 0.5),
      axis.title.y    = element_text(color = "black"),
      axis.text.y     = element_text(color = "black"),
      legend.position = "none"
    )
```

#### Plot 2: Summed number of prescriptions by year.

```{r plotYearRxSum, fig.width = 20, fig.height = 12}
sdudAggYear %>%
  ggplot() +
  geom_col(
    aes(
      x    = FakeDate,
      y    = TotalRx,
      fill = FakeDate
    ),
    color = "black"
  ) +
  scale_x_date(
    date_breaks = "1 year",
    date_labels = "%Y",
    expand      = c(.01, .01)
  ) +
  scale_y_continuous(labels = scales::comma) +
  scale_fill_viridis_c() + 
  theme_ipsum_rc(grid = "XY") +
  xlab("Year") +
  ylab("Number of Prescriptions (sum)") +
  labs(caption = "Source: https://www.medicaid.gov/medicaid/prescription-drugs/state-drug-utilization-data/index.html") +
  ggtitle("Summed Number of Prescriptions by Year - 2000 to 2020") +
  theme(
    axis.title.x    = element_text(color = "black"),
    axis.text.x     = element_text(color = "black", angle = 90, vjust = 0.5),
    axis.title.y    = element_text(color = "black"),
    axis.text.y     = element_text(color = "black"),
    legend.position = "none"
  )
```

#### Plot 3: Record counts by quarter and year.

```{r plotQtrYearRecordCount, fig.width = 20, fig.height = 12}
sdudAggQtrYear %>%
  ggplot() +
    geom_col(
      aes(
        x    = FakeDate,
        y    = RowCount,
        fill = factor(quarter)
      ),
    color = "black"
    ) +
    scale_x_date(
      date_breaks = "1 year",
      date_labels = "%Y",
      expand      = c(.01, .01)
    ) +
    scale_y_continuous(labels = scales::comma) +
    scale_fill_viridis_d() + 
    theme_ipsum_rc(grid = "XY") +
    xlab("Year") +
    ylab("Number of Records") +
    labs(
      caption = "Source: https://www.medicaid.gov/medicaid/prescription-drugs/state-drug-utilization-data/index.html",
      fill = "Quarter"
    ) +
    ggtitle("Number of Records by Quarter and Year - 2000 to 2020") +
    theme(
      axis.title.x    = element_text(color = "black"),
      axis.text.x     = element_text(color = "black", angle = 90, vjust = 0.5),
      axis.title.y    = element_text(color = "black"),
      axis.text.y     = element_text(color = "black")
    )
```

#### Plot 4: Summed number of prescriptions by quarter and year.

```{r plotQtrYearRxSum, fig.width = 20, fig.height = 12}
sdudAggQtrYear %>%
  ggplot() +
    geom_col(
      aes(
        x    = FakeDate,
        y    = TotalRx,
        fill = factor(quarter)
      ),
    color = "black"
    ) +
    scale_x_date(
      date_breaks = "1 year",
      date_labels = "%Y",
      expand      = c(.01, .01)
    ) +
    scale_y_continuous(labels = scales::comma) +
    scale_fill_viridis_d() + 
    theme_ipsum_rc(grid = "XY") +
    xlab("Year") +
    ylab("Number of Prescriptions (sum)") +
    labs(
      caption = "Source: https://www.medicaid.gov/medicaid/prescription-drugs/state-drug-utilization-data/index.html",
      fill = "Quarter"
    ) +
    ggtitle("Summed Number of Prescriptions by Quarter and Year - 2000 to 2020") +
    theme(
      axis.title.x    = element_text(color = "black"),
      axis.text.x     = element_text(color = "black", angle = 90, vjust = 0.5),
      axis.title.y    = element_text(color = "black"),
      axis.text.y     = element_text(color = "black")
    )
```

#### Plot 5: Number of records by state and year.

```{r plotStateYearRecordCount, fig.width = 20, fig.height = 12}
sdudAggStateYear %>%
  ggplot() +
    geom_col(
      aes(
        x    = FakeDate,
        y    = RowCount,
        fill = FakeDate
      ),
    color = "black"
    ) +
    scale_x_date(
      date_breaks = "1 year",
      date_labels = "%Y",
      expand      = c(.01, .01)
    ) +
    scale_y_continuous(labels = scales::comma) +
    scale_fill_viridis_c() + 
    facet_wrap(~state) +
    theme_ipsum_rc(grid = "XY") +
    xlab("Year") +
    ylab("Number of Records") +
    labs(caption = "Source: https://www.medicaid.gov/medicaid/prescription-drugs/state-drug-utilization-data/index.html") +
    ggtitle("Number of Records by State and Year - 2000 to 2020") +
    theme(
      axis.title.x    = element_text(color = "black"),
      axis.text.x     = element_text(color = "black", angle = 90, vjust = 0.5, size = 8),
      axis.title.y    = element_text(color = "black"),
      axis.text.y     = element_text(color = "black"),
      legend.position = "none"
    )
```

#### Plot 6: Summed number of prescriptions by state and year.

```{r plotStateYearRxSum, fig.width = 20, fig.height = 12}
sdudAggStateYear %>%
  ggplot() +
    geom_col(
      aes(
        x    = FakeDate,
        y    = TotalRx,
        fill = FakeDate
      ),
    color = "black"
    ) +
    scale_x_date(
      date_breaks = "1 year",
      date_labels = "%Y",
      expand      = c(.01, .01)
    ) +
    scale_y_continuous(labels = scales::comma) +
    scale_fill_viridis_c() + 
    facet_wrap(~state) +
    theme_ipsum_rc(grid = "XY") +
    xlab("Year") +
    ylab("Number of Prescriptions (sum)") +
    labs(caption = "Source: https://www.medicaid.gov/medicaid/prescription-drugs/state-drug-utilization-data/index.html") +
    ggtitle("Summed Number of Prescriptions by State and Year - 2000 to 2020") +
    theme(
      axis.title.x    = element_text(color = "black"),
      axis.text.x     = element_text(color = "black", angle = 90, vjust = 0.5, size = 8),
      axis.title.y    = element_text(color = "black"),
      axis.text.y     = element_text(color = "black"),
      legend.position = "none"
    ) 
```

#### Plot 7: Remove outliers and re-plot.

```{r plotStateYearRxSumRmOutliers, fig.width = 20, fig.height = 12}
sdudAggStateYear %>%
  filter(
    !(
      (state == "WA" & year == 2006) |
        (state == "SD" & year == 2007) |
        (state == "TN") & year %in% c(2003:2006)
    )
  ) %>%
  ggplot() +
    geom_col(
      aes(
        x    = FakeDate,
        y    = TotalRx,
        fill = FakeDate
      ),
    color = "black"
    ) +
    scale_x_date(
      date_breaks = "1 year",
      date_labels = "%Y",
      expand      = c(.01, .01)
    ) +
    scale_y_continuous(labels = scales::comma) +
    scale_fill_viridis_c() + 
    facet_wrap(~state) +
    theme_ipsum_rc(grid = "XY") +
    xlab("Year") +
    ylab("Number of Prescriptions (sum)") +
    labs(caption = "Source: https://www.medicaid.gov/medicaid/prescription-drugs/state-drug-utilization-data/index.html \n Outliers identified (WA                     2006, SD 2007, TN 2003 - 2006)") +
    ggtitle("Summed Number of Prescriptions by State and Year with Outliers Removed - 2000 to 2020") +
    theme(
      axis.title.x    = element_text(color = "black"),
      axis.text.x     = element_text(color = "black", angle = 90, vjust = 0.5, size = 8),
      axis.title.y    = element_text(color = "black"),
      axis.text.y     = element_text(color = "black"),
      legend.position = "none"
    )
```

#### Plot 8: Number of records by quarter, state, and year.

```{r plotQtrStateYearRecordCount, fig.width = 20, fig.height = 12}
sdudAggQtrStateYear %>%
  ggplot() +
    geom_col(
      aes(
        x    = FakeDate,
        y    = RowCount,
        fill = factor(quarter)
      ),
      color = "black"
    ) +
    scale_x_date(
      date_breaks = "1 year",
      date_labels = "%Y",
      expand      = c(.01, .01)
    ) +
    scale_y_continuous(labels = scales::comma) +
    scale_fill_viridis_d() + 
    facet_wrap(~state) +
    theme_ipsum_rc(grid = "XY") +
    xlab("Year") +
    ylab("Number of Records") +
    labs(
      caption = "Source: https://www.medicaid.gov/medicaid/prescription-drugs/state-drug-utilization-data/index.html",
      fill = "Quarter"
    ) +
    ggtitle("Number of Records by Quarter, State, and Year - 2000 to 2020") +
    theme(
      axis.title.x    = element_text(color = "black"),
      axis.text.x     = element_text(color = "black", angle = 90, vjust = 0.5, size = 8),
      axis.title.y    = element_text(color = "black"),
      axis.text.y     = element_text(color = "black")
    )
```

#### Plot 9: Summed number of prescriptions by quarter, state, and year.

```{r plotQtrStateYearRxSum, fig.width = 20, fig.height = 12}
sdudAggQtrStateYear %>%
  ggplot() +
    geom_col(
      aes(
        x    = FakeDate,
        y    = TotalRx,
        fill = factor(quarter)
      ),
      color = "black"
    ) +
    scale_x_date(
      date_breaks = "1 year",
      date_labels = "%Y",
      expand      = c(.01, .01)
    ) +
    scale_y_continuous(labels = scales::comma) +
    scale_fill_viridis_d() + 
    facet_wrap(~state) +
    theme_ipsum_rc(grid = "XY") +
    xlab("Year") +
    ylab("Number of Prescriptions (sum)") +
    labs(
      caption = "Source: https://www.medicaid.gov/medicaid/prescription-drugs/state-drug-utilization-data/index.html",
      fill = "Quarter"
    ) +
    ggtitle("Summed Number of Prescriptions by Quarter, State, and Year - 2000 to 2020") +
    theme(
      axis.title.x    = element_text(color = "black"),
      axis.text.x     = element_text(color = "black", angle = 90, vjust = 0.5, size = 8),
      axis.title.y    = element_text(color = "black"),
      axis.text.y     = element_text(color = "black")
    )
```

#### Plot 10: Remove outliers and re-plot.

```{r plotQtrStateYearRxSumRmOutliers, fig.width = 20, fig.height = 12}
sdudAggQtrStateYear %>%
  filter(
    !(
      (state == "WA" & year == 2006 & quarter == 3) |
        (state == "SD" & year == 2007 & quarter == 3) |
        (state == "TN") & year %in% c(2003:2006)
    )
  ) %>%
  ggplot() +
    geom_col(
      aes(
        x    = FakeDate,
        y    = TotalRx,
        fill = factor(quarter)
      ),
      color = "black"
    ) +
    scale_x_date(
      date_breaks = "1 year",
      date_labels = "%Y",
      expand      = c(.01, .01)
    ) +
    scale_y_continuous(labels = scales::comma) +
    scale_fill_viridis_d() + 
    facet_wrap(~state) +
    theme_ipsum_rc(grid = "XY") +
    xlab("Year") +
    ylab("Number of Prescriptions (sum)") +
    labs(
      caption = "Source: https://www.medicaid.gov/medicaid/prescription-drugs/state-drug-utilization-data/index.html \n Outliers identified (WA                    2006 Q3, SD 2007 Q3, TN 2003-2006)",
      fill = "Quarter"
    ) +
    ggtitle("Summed Number of Prescriptions by Quarter, State, and Year with Outliers Removed - 2000 to 2020") +
    theme(
      axis.title.x    = element_text(color = "black"),
      axis.text.x     = element_text(color = "black", angle = 90, vjust = 0.5, size = 8),
      axis.title.y    = element_text(color = "black"),
      axis.text.y     = element_text(color = "black")
    ) 
```
